# Viewing The Data {-}

Before visualising any data set, we first need to know its contents. For example, the contents of the `flights` data within the `nycflights13` library can be observed using the following command:

```{r flights, echo = TRUE, eval = FALSE}
head(flights, n = 3)
```

This prints to the R console the first `n = 3` rows of the `flights` data set, displaying each of the variables within said data set. We now know the data set contains `r ncol(flights)` variables, as well as their names. A quick check on the size of a data set can be obtained using:

```{r flights2, echo = TRUE, eval = TRUE}
dim(flights)
```

which displays the dimensions of the data set. Thus, here we have `r nrow(flights)` rows and `r ncol(flights)` columns worth of data.

To reduce the amount of data we will be working with and make things a little easier, let's only look at *Alaska Airlines* flights leaving from New York City in 2013. This can be done by subsetting the data in such a way that we only observe flights from Alaska Airlines (carrier code AS), as follows:

```{r alaska, echo = TRUE, eval = TRUE}
Alaska <- flights[flights$carrier == "AS", ]
```

This essentially picks out all of the rows within the `flights` data set for which the carrier code is `AS` and discards the rest, thus creating a new data set entitled `Alaska`.

<br>

**Task**: Write code to observe the first 5 rows of the `Alaska` data.

`r hide("Hint")`
You may want to use the `head` function.
`r unhide()`

```{r Task1-solution, webex.hide="Solution", echo=TRUE, eval=TRUE, warning=FALSE}
head(Alaska, n = 5)
```

```{r MCQ1}
opts_Q1 <- sample(c("500 19",
                    "19 500",
                    answer = "714 19",
                    "19 714"))
```

<br>

**What are the dimensions of the `Alaska` data set?**
`r longmcq(opts_Q1)`

`r hide("Hint")`
Check the the dimensions using the `dim` function.
`r unhide()`

<br>

Next week we will look at more sophisticated ways of manipulating data sets. Now, let us go on to look at different visualisations of our `Alaska` data set using `ggplot2`, starting with **scatterplots**.

<br>
<br>

